---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# ggnested
`ggnested` is a wrapper function around `ggplot2` that enables users to display data 
with a two-level nested or clustered structure. It uses colours or fills to label the main group
or cluster, and a gradient of shades and tints of each colour to label subgroups within each 
main group. It is particularly useful for data with a hierarchy, such as taxonomic annotations or
geographic locations. 

This is one of the core functions of the `fantaxtic` package, which
visualizes taxonomic ranks of amplicon sequencing / microbiome data as stored in 
`phyloseq` objects. However, as it has utility beyond displaying taxonomic annotations,
it has been implemented as a separate package.

**Keywords: R, ggplot2, nested data, clustered data, multiple levels, shades, tints, gradient**

## Installation
```{r, eval = FALSE}
if(!"devtools" %in% installed.packages()){
  install.packages("devtools")
}
devtools::install_github("gmteunisse/ggnested")
```

## Usage
### Basic
`ggnested` can be used exactly like the regular `ggplot` function, but with some extra arguments. Instead of specifying a color or a fill, a `main_group` and a `sub_group` need to be specified in the aesthetic mapping through `aes(...)`. The ggplot object can then be used to create any regular ggplot plot, such as barplots, scatterplots or boxplots, as well as any other layers such as faceting.

```{r load_package, results='hide', message=F, warning=F}
library(ggnested)
data(diamonds)
```

```{r barplot}
ggnested(diamonds, 
         aes(clarity, 
             main_group = color, 
             sub_group = cut)) + 
  geom_bar()
``` 

```{r scatterplot, fig.width = 12}
ggnested(diamonds, 
         aes(x = carat, 
             y = price, 
             main_group = color, 
             sub_group = cut)) + 
  geom_point(alpha = 0.5) +
  facet_wrap(~color)
```

```{r boxplot}
ggnested(diamonds, 
         aes(x = color,
             y = price, 
             main_group = color, 
             sub_group = cut)) + 
  geom_boxplot(alpha = .5)

```

### Options
A few options are available, most of which are related to the appearance of the plot.

#### Legend
A few options are available to alter the way the legend is constructed. 

* The `legend_labeling` option alters the labels that are used in the legend. For example, it is possible to display both the `main_group` label and the `sub_group` label, or just the `main_group` label.
* The `join_str` option is used if `legend_labeling = "join"`,and specifies a string to join the two labels.
* The `legend_title` option alters the legend title.
* The `main_keys` determines whether `main_group` labels are displayed as keys within the legend.

Bringing it all together:

```{r legend_labeling}
ggnested(diamonds, 
         aes(clarity, 
             main_group = color, 
             sub_group = cut),
         legend_labeling = "join",
         join_str = ": ",
         legend_title = "Joined labels without titles",
         main_keys = FALSE) + 
  geom_bar()

```

#### Colour palette
`ggnested` creates its own colour palettes. Like ggplot2, it samples maximally separated colours, but unlike ggplot2, it shuffles the palette so that dissimlar colours are placed next to one another. This is to avoid confusion in colours and shades between different groups. Also unlike ggplot2, the user can define a base colour (e.g. `base_clr = #008CF0`) that serves as the first colour of the palette, based on which all other colours are sampled. Finally, A custom palette of colours for the `main_group` can be provided (`main_palette = c()`), which will then be used to generate gradients for the subgroups.

```{r base_clr}
ggnested(diamonds, 
         aes(clarity, 
             main_group = color, 
             sub_group = cut),
         base_clr = "red") + 
  geom_bar()

```

Note that if a custom palette is named, the colours will be associated with the appropriate `main_group` level.

```{r}
pal <- c("red", "blue", "green", "yellow", "purple", "orange", "grey50")
names(pal) <- unique(diamonds$color)
ggnested(diamonds, 
         aes(clarity, 
             main_group = color, 
             sub_group = cut), 
         main_palette = pal) + 
  geom_bar()


```

#### Affected aethetics
By default, both the `color` and `fill` aesthetics of the plot are altered by `ggnested`. However, it is possible to specify which aesthetics should be altered using `nested_aes = c("fill", "colour")`. Note that it leads to issues with the legends of geoms that automatically create both `colour` and `fill` aethetics, as shown with the boxplots below - the `main_group` titles now also show a box outline! This can of course be resolved by setting `main_keys = FALSE` as shown above.
```{r nested_aes}
ggnested(diamonds, 
         aes(x = color,
             y = price, 
             main_group = color, 
             sub_group = cut),
         nested_aes = "fill") + 
  geom_boxplot()

```

### Themes
To ensure that a `ggnested` plot does not alter when the theme is altered, it is recommended to alter the theme using the `theme_nested` function. This function is simply a wrapper around any ggplot `theme_*` function.

```{r}
ggnested(diamonds, 
         aes(clarity, 
             main_group = color, 
             sub_group = cut)) + 
  geom_bar() +
  theme_nested(theme_minimal)

```
